---
title: "Some Examples Using Simulation"
author: "Ryan Lafferty"
date: "2/28/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all(path=r"(C:\Users\laffertyrm\Documents\work\PB\pbanalysis)")

```

### Simulated Racial Discrimination 

 Let's try to simulate a situation where discrimination is happening 
 and see if the PB method can detect it.
 Suppose black and white applicants are applying for a certain job.
 Imagine that it is permissible to judge applicants based on educational 
 background, years of experience in the role, and results from a certain aptitude
 test administered to all applicants. However, we assume that it is not permissible
 to judge an applicant directly on the basis of his/her race. 
 
 Let us define the three covariates of interest. The first will be a binary 
 measure of educational background, `college` which indicates whether the applicant 
 graduated college or not. The second, `experience` will be the number of years
 the applicant has worked in a related role. The third, `score` is the the 
 applicant's test score on the aptitude test, which we will suppose is a number
 between 0 and 100. 
 
 We will assume that the first two covariates may depend on the applicant's race,
 due to possible socio-economic factors, while the third has no relationship to
 the applicant's race. 
 
 ```{r}
# Define the race category, and the three covariates

n_white = 500
n_black = 500
n = n_white + n_black

race = rep("black", n)
race[1:n_white] = "white"

# bernoulli with higher success probability for whites
college = rep(0,n)
college[race == "white"] = rbinom(n_white,size = 1, prob = .44)
college[race == "black"] = rbinom(n_black,size = 1, prob = .30)

# exponential with higher mean for whites
experience = rep(0,n)
experience[race == "white"] = round(rexp(n_white,rate=1/2))
experience[race == "black"] = round(rexp(n_black,rate=1/1))

# approx normally distributed but cut off at 0 and 100
score = round(pmax(0,pmin(100,rnorm(n,mean=70,sd=15))),2)

 ```
 
 
 The first case we will consider is the honest employer which does not take race
 into account in its application process. Let's call this Employer A. The 
 response variable, `hired` will be a binary variable that records whether an
 applicant has been hired or not following the application process. We will 
 imagine that Employer A makes its hiring decisions using the following
 idealized method: First, we randomly generate some (positive) coefficients for 
 each covariate (not including race). Then, we multiply the (normalized) covariates
 by the corresponding coefficients and sum for each applicant. Then we hire the 
 top ten applicants with the highest totals and reject the rest.
 
 ```{r}
 # First randomly generate some positive coefficients for hiring function
 coefs = runif(3)
 
 # Normalize the covariates 
 norm_college = (college-mean(college))/sd(college)
 norm_experience = (experience-mean(experience))/sd(college) 
 norm_score = (score - mean(score))/sd(score)
 
 # Compute weighted totals
 totals = coefs[1]*norm_college + coefs[2]*norm_experience + coefs[3]*norm_score
 totals[race == "black"] = totals[race == "black"]
 
 # Choose top ten totals and define response variable
 top.10.idx = order(totals,decreasing=T)[1:10]
 hired = rep("rejected",n)
 hired[top.10.idx] = "hired"
 
 ```
 
 Now let us test our pb.fit function on this simulated scenario. We will assume
 a binary logistic model. 
 
 ```{r}
 data = data.frame(hired = hired,
                   race = race,
                   college = college,
                   experience = experience,
                   score = score)
 results = pb.fit(hired ~ college + experience + score,
                  data = data,
                  family = "multinomial",
                  disparity.group = "race",
                  majority.group = "white",
                  minority.group = "black",
                  base.level = "hired")
 print(results)
 ```
